---
title: "Cyclistic Bike-Share Case Study"
author: "Eduardo Carrascosa"
date: "2/21/2022"
output: 
bookdown :: html_document2: default
tidy: styler
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  )

# remove scientific notiation
options(scipen = 100)
```

```{r install-libraries}
```


```{r install-libraries}
library(tidyverse)
library(DT)
library(caret)
library(lattice)
library(lubridate)
library(scales)
library(ggthemes)

```
# Case Study: How Does a Bike-Share Navigate Speedy Success?
You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.


## PHASE 1: Ask
Three questions will guide the future marketing program:
1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

### Business Task
Use the data to identify ways the ways that casual riders use bikes differently than members and recommend ways to convert casual riders to members.

## PHASE 2: Prepare
You will use Cyclistic’s historical trip data to analyze and identify trends. [Download the previous 12 months of Cyclistic trip data here.](https://divvy-tripdata.s3.amazonaws.com/index.html) (Note: The datasets have a different name because Cyclistic is a fictional company. For the purposes of this case study, the datasets are appropriate and will enable you to answer the business questions. The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement.) This is public data that you can use to explore how different customer types are using Cyclistic bikes. But note that data-privacy issues prohibit you from using riders’ personally identifiable information. This means that you won’t be able to connect pass purchases to credit card numbers to determine if casual riders live in the Cyclistic service area or if they have purchased multiple single passes.

### Key tasks 
1. Download data and store it appropriately. 
2. Identify how it’s organized. 
3. Sort and filter the data.

```{r load-data}
oct_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202010-divvy-tripdata.csv")
nov_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202011-divvy-tripdata.csv")
dec_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202012-divvy-tripdata.csv")
jan_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202101-divvy-tripdata.csv")
feb_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202102-divvy-tripdata.csv")
mar_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202103-divvy-tripdata.csv")
apr_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202104-divvy-tripdata.csv")
may_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202105-divvy-tripdata.csv")
jun_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202106-divvy-tripdata.csv")
jul_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202107-divvy-tripdata.csv")
aug_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202108-divvy-tripdata.csv")
sep_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202109-divvy-tripdata.csv")

```
## Peek at the datasets {.tabset}

### Oct 2020
```{r results='asis'}
datatable(head(oct_2020, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

### Nov 2020
```{r results='asis'}
datatable(head(nov_2020, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

### Dec 2020
```{r results='asis'}
datatable(head(dec_2020, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### Jan 2021
```{r results='asis'}
datatable(head(jan_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

### Feb 2021
```{r results='asis'}
datatable(head(feb_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

### Mar 2021
```{r results='asis'}
datatable(head(mar_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### Apr 2021
```{r results='asis'}
datatable(head(apr_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### May 2021
```{r results='asis'}
datatable(head(may_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### Jun 2021
```{r results='asis'}
datatable(head(jun_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### Jul 2021
```{r results='asis'}
datatable(head(jul_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

### Aug 2021
```{r results='asis'}
datatable(head(aug_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))
```

### Sep 2021
```{r results='asis'}
datatable(head(sep_2021, 100), style="bootstrap", class="table-condensed", options = list(dom = 'tp',scrollX = TRUE))

```

## WRANGLE DATA AND COMBINE INTO A SINGLE FILE

We need to check that all column names match in order to combine all the data into a single file. 


```{r column-name-check, results='hide'}
colnames(oct_2020) 
colnames(nov_2020)
colnames(dec_2020)
colnames(jan_2021)
colnames(feb_2021)
colnames(mar_2021)
colnames(apr_2021)
colnames(may_2021)
colnames(jun_2021)
colnames(jul_2021)
colnames(aug_2021)
colnames(sep_2021)
```

All column names match, now let's peak at the dataset and look for incongruencies.

```{r glimpse, results='hide'}
glimpse(oct_2020)
glimpse(nov_2020)
glimpse(dec_2020)
glimpse(jan_2021)
glimpse(feb_2021)
glimpse(mar_2021)
glimpse(apr_2021)
glimpse(may_2021)
glimpse(jun_2021)
glimpse(jul_2021)
glimpse(aug_2021)
glimpse(sep_2021)
```

Convert start/end times to datetime to match other files and let's combine the datasets in one

```{r combine-datasets}

oct_2020 <- oct_2020 %>% 
              mutate(started_at = mdy_hm(started_at))

oct_2020 <- oct_2020 %>%
                mutate(ended_at = mdy_hm(ended_at))

# Convert station id to character to match other files
oct_2020 <- oct_2020 %>%
              mutate(start_station_id = as.character(start_station_id))

oct_2020 <- oct_2020 %>%
              mutate(end_station_id = as.character(end_station_id))

# Convert start_station_id to character to match other files
nov_2020 <- mutate(nov_2020, start_station_id = as.character(start_station_id))

# Convert end_station_id to character to match other files
nov_2020 <- mutate(nov_2020, end_station_id = as.character(end_station_id))


# Stack all monthly files into a dataframe
all_trips <- bind_rows(oct_2020, nov_2020, dec_2020, jan_2021, feb_2021, mar_2021,
                        apr_2021, may_2021, jun_2021, jul_2021, aug_2021, sep_2021)
```

Now let's get a glimpse of the data file:
```{r glimpse-entire-file}
glimpse(all_trips)
```

## INSPECT DATA AND PREPARE FOR ANALYSIS
```{r}
colnames(all_trips) # List of column names
nrow(all_trips) # How many rows are in the data frame?
dim(all_trips) # Dimensions of the data frame?
head(all_trips) # See the first 6 rows of data frame. 
tail(all_trips) # See last 6 rows of data frame.
str(all_trips) # See list of columns and data types.
summary(all_trips) # Statistical summary of data.
glimpse(all_trips) # Summary
```

### Drop unused data
```{r}
all_trips <- all_trips %>%
                select(-end_station_id)
```

### Check for null values

```{r}
colSums(is.na(all_trips))
```

### Handle missing data

```{r}
all_trips <- all_trips %>% drop_na()
```

Handle dates
```{r add-date-data }
all_trips$start_date <- as.Date(all_trips$started_at)

all_trips$stop_date <- as.Date(all_trips$ended_at)

all_trips$month <- format(as.Date(all_trips$start_date), '%m')

all_trips$day <- format(as.Date(all_trips$start_date), '%d')

all_trips$year <- format(as.Date(all_trips$start_date), '%Y')

all_trips$day_of_week <- format(as.Date(all_trips$start_date), '%A')

```

### Add ride length and convert to numeric

```{r ride-length}
all_trips <- mutate(all_trips, ride_length = difftime(all_trips$ended_at, all_trips$started_at))

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```
```{r}
is.numeric(all_trips$ride_length)
```

Create new dataset, remove bad data

```{r}
all_trips_v2 <- all_trips %>%
    filter(ride_length > 0)
```

Convert ride length to minutes
```{r}
all_trips_v2 <- all_trips_v2 %>%
  mutate(ride_length_in_minutes = ride_length/60)
```

 Most frequent start stations

```{r most-popular-station}

all_trips_v2 %>%
  group_by(start_station_name) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  slice_max(count, n = 20) %>%
  mutate(name = fct_reorder(start_station_name, count)) %>%
  ggplot(aes(x = name, y = count, fill = name)) +
  geom_col()+
  scale_fill_hue(c = 40) +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")
```

```{r most-popular-start-date}
all_trips_v2 %>%
  group_by(start_date) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  slice_max(count, n = 20) %>%
  mutate(date = fct_reorder(as.factor(start_date), count)) %>%
  ggplot(aes(x = date, y = count, fill = date)) +
  geom_col()+
  scale_fill_hue(c = 40) +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")
```
```{r popular-stop-date}
all_trips_v2 %>%
  group_by(stop_date) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  slice_max(count, n = 20) %>%
  mutate(date = fct_reorder(as.factor(stop_date), count)) %>%
  ggplot(aes(x = date, y = count, fill = date)) +
  geom_col()+
  scale_fill_hue(c = 40) +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")
```

User Type

```{r}
  # Plot casual vs member count
ggplot(all_trips) + 
  geom_bar(mapping = aes(x = member_casual, fill = member_casual)) +
  theme_bw() +
  theme(legend.position="none")

```

Trip duration

```{r}
# Visual 3: facet grid showing distribution of rides per bike type
###########################
  
all_trips_v2 %>%
  group_by(month, member_casual, rideable_type) %>%
  filter(member_casual == 'casual') %>%
  summarise(
    number_of_rides = n(), 
    average_duration = mean(ride_length)
  ) %>%
  ggplot( aes(x = month, y = average_duration, color = rideable_type) ) +
  geom_point(size = 3) +
  facet_wrap(~rideable_type)
```
ˆ
```{r}
# Visual 4: Facet grid showing casual vs members dist by bike type
############################################

all_trips_v2 %>%
  group_by(month, member_casual, rideable_type) %>%
  summarise(
    number_of_rides = n(), 
    average_duration = mean(ride_length)
  ) %>%
  ggplot( aes(x = month, y = number_of_rides, color = rideable_type) ) +
  geom_point(size = 3) +
  facet_grid(rideable_type ~ member_casual)
```
```{r}
#############################################
# Visual 5: Number of rides per month
############################################
all_trips_v2 %>%
  group_by(member_casual, month) %>%
  summarise(
    number_of_rides = n()
  ) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
```{r}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    average_duration = mean(ride_length)
  ) %>%
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```
```{r}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(ride_length = mean(ride_length)) %>%
  ggplot(aes(x = day_of_week, y = ride_length, fill = member_casual)) +
  geom_col(position = "stack")
```

```{r}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    avg_ride_length = mean(ride_length), 
    number_of_rides = n()
  ) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge")
```


