---
title: "Recency, Frequency, Monetary (RFM) Analysis"
author: "Eduardo Carrascosa"
date: "2/2/2022"
output: 
  html_document:
    fig_height: 4
    fig_width: 7
    theme: cosmo
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  )
  
# remove scientific notiation
options(scipen = 100)
```

# Recenty-Frequency-Monetary (RFM) Analysis
______

Upscale specialty catalog retailer with a mix of “staple” items and seasonal, trendier items, using accumulated purchase history to increase the effectiveness and efficiency of marketing expenditures.

Recency: how long ago the customer last made a purchase

Frequency: how many purchases the customer has made  

Monetary: how much each customer has spent in total  


## PHASE 1: Business Task

 1. Conduct RFM analysis to assess the effectiveness of a customer targeting performance
 
 2. Draw cummulative profit, lift, and gains charts
 
## PHASE 2: Prepare

### Key tasks 

1. Download data and store it appropriately. 
2. Identify how it’s organized. 
3. Sort and filter the data.

```{r load-data}
library(tidyverse) # metapackage of all tidyverse packages

# Input data files are available in the read-only "../input/" directory
# For example, running this (by clicking run or pressing Shift+Enter) will list all files under the input directory


# load data
RFM <- read_csv("rfm.csv")
```

##  PHASE 3: Process

### Key tasks
1. Calculate total number of customers who receive the catalogs and the number of buyers
2. Calculate the average response rate
3. Calculate the response rate, lift, gain, and profit per cell

```{r }
# parameters of the problem
avg_order <- 104.24
pr_margin <- 0.5
var_cost <- 1

# Calculate total number of customers who receive the catalogs
total_customers <- sum(RFM$customers)

# Calculate total number of customers who ordered from the catalogs
total_buyers <- sum((RFM$buyers))

# Calculate the average response rate
avg_rr <- total_buyers / total_customers

# Calculate the response rate, lift, gain, and profit per cell
RFM <- mutate(RFM, response_rate = buyers / customers )

RFM <- mutate(RFM, lift = response_rate / avg_rr * 100 )

RFM <- mutate(RFM, gain = buyers / total_buyers * 100)

RFM <- mutate(RFM, profit = buyers * avg_order * pr_margin - var_cost * customers) # complete the statement

total_profit <- sum(RFM$profit)
```

## PHASE 4: Analyze

### Key Tasks

1. What is the profit, lift, and gain if we only target the cells with response rate > 1.94?

2. Calculate the cumulative lift, gain, and profit per cell in order to evaluate the performance of targeting.

The average response rate if 2.45%, if we only target customers with a response rate 
of 1.94% and above our response rate increases to 3.35%. Similary If we only target 
the cells with a reponse rate of 1.94% or higher our profit increases by $12,088.72. 
The lift for this group as measured by dividing the response rate by the average 
response rate is 36% and we gain 73 customers.

```{r }
# check for number of obs
RFM %>%
  filter(response_rate > .0194) %>%
  summarise(count = n())

# set with response rate greater than 1.94%
RR_GT_1_94 <- filter(RFM, response_rate > .0194) 

# profit for the set, $3 less than max profit
profit_1_94 <-sum(RR_GT_1_94$profit)

# response rate for this group is 3.35 %
rr_1_94 <- sum(RR_GT_1_94$buyers) / sum(RR_GT_1_94$customers)

# lift for this group is 136
lift_1_94 <- rr_1_94 / avg_rr * 100

# gain is 73.89
gain_1_94 <- sum(RR_GT_1_94$buyers) / total_buyers * 100

# Calculate the cumulative lift, gain, and profit per cell in order to evaluate the performance of targeting

RFM <- arrange(RFM, desc(response_rate)) # start by ranking the customers by response rate

RFM <- mutate(RFM, cumsum_customer = cumsum(customers)) # cumsum customers

RFM <- mutate(RFM, cumsum_buyers = cumsum(buyers) ) # cumsum buyers

RFM <- mutate(RFM, customer_percent_rr = cumsum(customers) / total_customers * 100)

RFM <- mutate(RFM, cumm_response_rate = cumsum(buyers) / cumsum(customers)) # allows targeting by response rate percentile

RFM <- mutate(RFM, cum_lift_rr = cumsum(buyers) / cumsum(customers) / avg_rr * 100)

RFM <- mutate(RFM, cum_gain_rr = cumsum(buyers) / total_buyers * 100)

RFM <- mutate(RFM, cum_profit = cumsum(profit))

RFM <- mutate(RFM, profit_percentile = cum_profit / max(cum_profit) * 100)
```

## PHASE 5: Share

###  Key Tasks:
 1. Determine the best way to share your findings.
 2. Create effective data visualizations. 
 3. Present your findings. 
 4. Ensure your work is accessible.

```{r}

 # Plot the charts

 ggplot(RFM) + geom_line(aes(x=customer_percent_rr, y=cum_profit))
## This graph shows how much profit you capture by targeting different percentiles of the sample group. 
## If I target top 25 percentile of customers ranked by avg response rate I capture over 30K in profits. 
## If I target the 50 percentile of customers I maximize my profit at close to 40K
## The goal is NOT ALWAYS maximize profit. Sometimes the goal may be to increase
## market share. In that case I sacrifice some profit to increase market
## share by going for a larger percentile even when the profit curve
## goes down.

ggplot(RFM) + geom_line(mapping = aes(x = customer_percent_rr, y = profit_percentile ))


ggplot(RFM) + geom_line(aes(x=customer_percent_rr, y=cum_lift_rr))

ggplot(RFM) + geom_line(aes(x=customer_percent_rr, y=cum_gain_rr))
## If I target top 50% I get 75% of the customer that I get if I target the entire 95K sample.
## If I target top 25 percentile of customers I get close to 50% of all the customers I get if
## I target the entire 95K


```

## Key Takeaways

* The top 10% of customers make 51% of profits, top 25% make 80% of profits
* If we target the top 50% of customers we get 75% of the buyers we would get if we target the entire sample (gain)
* If we target the top 10% of customers by response rate our response rate will be 5.83% and we capture 23% of the buyers (gain) that we would get if we target the entire sample.
* If we target the top 10% percentile our lift in buyers is 2.37 times the sample.
* Profit starts to decline at the 50th percentile 