---
title: "Cyclistic Bike-Share Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Eduardo Carrascosa"
date: "2/21/2022"
output: 
  html_document:
    fig_height: 4
    fig_width: 7
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  )

# remove scientific notiation
options(scipen = 100)
```

```{r install-libraries}
library(tidyverse)
library(DT)
library(caret)
library(lattice)
library(lubridate)
library(scales)
library(ggthemes)

```

```{r load-data}
oct_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202010-divvy-tripdata.csv")
nov_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202011-divvy-tripdata.csv")
dec_2020 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202012-divvy-tripdata.csv")
jan_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202101-divvy-tripdata.csv")
feb_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202102-divvy-tripdata.csv")
mar_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202103-divvy-tripdata.csv")
apr_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202104-divvy-tripdata.csv")
may_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202105-divvy-tripdata.csv")
jun_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202106-divvy-tripdata.csv")
jul_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202107-divvy-tripdata.csv")
aug_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202108-divvy-tripdata.csv")
sep_2021 <- read_csv("/Users/eduardocarrascosa/Desktop/Coursera/Capstone/bike_data/202109-divvy-tripdata.csv")


```
```{r column-name-check, results='hide'}
colnames(oct_2020) 
colnames(nov_2020)
colnames(dec_2020)
colnames(jan_2021)
colnames(feb_2021)
colnames(mar_2021)
colnames(apr_2021)
colnames(may_2021)
colnames(jun_2021)
colnames(jul_2021)
colnames(aug_2021)
colnames(sep_2021)
```

# Overview

In this project I will share my findings, including visualizations, key takeaways, and reccomendations
from the Cyclistic Bike Share case study project that was my capstone project during Google's Coursera Data Analytics Certificate.
For the purposes of this page I skip the earlier phases of the project and jump straight to sharing
my insights and visualizations. 


# Data visualizations
```{r glimpse, results='hide'}
glimpse(oct_2020)
glimpse(nov_2020)
glimpse(dec_2020)
glimpse(jan_2021)
glimpse(feb_2021)
glimpse(mar_2021)
glimpse(apr_2021)
glimpse(may_2021)
glimpse(jun_2021)
glimpse(jul_2021)
glimpse(aug_2021)
glimpse(sep_2021)
```




```{r combine-datasets}

oct_2020 <- oct_2020 %>% 
              mutate(started_at = mdy_hm(started_at))

oct_2020 <- oct_2020 %>%
                mutate(ended_at = mdy_hm(ended_at))

# Convert station id to character to match other files
oct_2020 <- oct_2020 %>%
              mutate(start_station_id = as.character(start_station_id))

oct_2020 <- oct_2020 %>%
              mutate(end_station_id = as.character(end_station_id))

# Convert start_station_id to character to match other files
nov_2020 <- mutate(nov_2020, start_station_id = as.character(start_station_id))

# Convert end_station_id to character to match other files
nov_2020 <- mutate(nov_2020, end_station_id = as.character(end_station_id))


# Stack all monthly files into a dataframe
all_trips <- bind_rows(oct_2020, nov_2020, dec_2020, jan_2021, feb_2021, mar_2021,
                        apr_2021, may_2021, jun_2021, jul_2021, aug_2021, sep_2021)
```




```{r results = 'hide'}
colnames(all_trips) # List of column names
nrow(all_trips) # How many rows are in the data frame?
dim(all_trips) # Dimensions of the data frame?
head(all_trips) # See the first 6 rows of data frame. 
tail(all_trips) # See last 6 rows of data frame.
str(all_trips) # See list of columns and data types.
summary(all_trips) # Statistical summary of data.
glimpse(all_trips) # Summary
```

```{r}
all_trips <- all_trips %>%
                select(-end_station_id)
```


```{r results='hide'}
colSums(is.na(all_trips))
```



```{r}
all_trips <- all_trips %>% drop_na()
```



```{r add-date-data }
all_trips$start_date <- as.Date(all_trips$started_at)

all_trips$stop_date <- as.Date(all_trips$ended_at)

all_trips$month <- format(as.Date(all_trips$start_date), '%m')

all_trips$day <- format(as.Date(all_trips$start_date), '%d')

all_trips$year <- format(as.Date(all_trips$start_date), '%Y')

all_trips$day_of_week <- format(as.Date(all_trips$start_date), '%A')

```

```{r ride-length}
all_trips <- mutate(all_trips, ride_length = difftime(all_trips$ended_at, all_trips$started_at))

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```


```{r}
all_trips_v2 <- all_trips %>%
    filter(ride_length > 0)
```

```{r}
all_trips_v2 <- all_trips_v2 %>%
  mutate(ride_length_in_minutes = ride_length/60)
```



```{r results = 'hide'}
# Descriptive analysis on ride_length (in seconds)
mean(all_trips_v2$ride_length) # Straight average (total ride length / rides )
median(all_trips_v2$ride_length) # Mid point ride length in sorted array of ride lengths
max(all_trips_v2$ride_length) # Longest ride
min(all_trips_v2$ride_length) # Shortest ride
```


```{r results = 'hide'}
summary(all_trips_v2$ride_length)
```


```{r, results = 'hide'}
#Group stats: max, min, mean, median by group
all_trips_v2 %>%
  group_by(member_casual) %>%
  summarise(mean_ride_length = mean(ride_length),
            median_ride_length = median(ride_length),
            max_ride_length = max(ride_length),
            min_ride_length = min(ride_length))
```

```{r}
# Notice that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


```{r results = 'hide'}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    avg_ride_length = mean(ride_length), 
    number_of_rides = n()
  ) 
```

```{r results = 'hide'}
# Count members and casual riders
all_trips %>%
  group_by(member_casual)%>%
  summarise(count = n())
```
### Casual users have longer ride durations on average than members
```{r}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    average_duration = mean(ride_length)
  ) %>%
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Casual riders use bikes mostly in the weekends
```{r}
all_trips_v2 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(
    avg_ride_length = mean(ride_length), 
    number_of_rides = n()
  ) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge")
```

### Streeter Dr & Grand Ave was the most popular bike station

```{r most-popular-station}

all_trips_v2 %>%
  group_by(start_station_name) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  slice_max(count, n = 20) %>%
  mutate(name = fct_reorder(start_station_name, count)) %>%
  ggplot(aes(x = name, y = count, fill = name)) +
  geom_col()+
  scale_fill_hue(c = 40) +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")
```

### The most popular start date was August 14th and 18 of the 20 busiest start dates occurred in the summer months
```{r most-popular-start-date}
all_trips_v2 %>%
  group_by(start_date) %>%
  summarise(count = n()) %>%
  arrange(-count) %>%
  slice_max(count, n = 20) %>%
  mutate(date = fct_reorder(as.factor(start_date), count)) %>%
  ggplot(aes(x = date, y = count, fill = date)) +
  geom_col()+
  scale_fill_hue(c = 40) +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")
```

### Docked bikes average the higher duration of the three types of bikes

```{r}
# Visual 3: facet grid showing distribution of rides per bike type
###########################
  
all_trips_v2 %>%
  group_by(month, member_casual, rideable_type) %>%
  filter(member_casual == 'casual') %>%
  summarise(
    number_of_rides = n(), 
    average_duration = mean(ride_length)
  ) %>%
  ggplot( aes(x = month, y = average_duration, color = rideable_type) ) +
  geom_point(size = 3) +
  facet_wrap(~rideable_type)
```

### Classic bikes are the most popular type of bikes among both groups
```{r}
# Visual 4: Facet grid showing casual vs members dist by bike type
############################################

all_trips_v2 %>%
  group_by(month, member_casual, rideable_type) %>%
  summarise(
    number_of_rides = n(), 
    average_duration = mean(ride_length)
  ) %>%
  ggplot( aes(x = month, y = number_of_rides, color = rideable_type) ) +
  geom_point(size = 3) +
  facet_grid(rideable_type ~ member_casual)
```

### The busiest months of the year are the summer months for both groups
```{r}
#############################################
# Visual 5: Number of rides per month
############################################
all_trips_v2 %>%
  group_by(member_casual, month) %>%
  summarise(
    number_of_rides = n()
  ) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```




# Key Takeaways

- Casual riders rent bikes mostly during the weekends
- The average duration per trip is significantly higher for casual riders than for members
- The number of rides per day for casual riders surpasses the number of rides per day for members during the summer months of June, July, and August probably due to tourism

# Recommendations

- Offer casual riders a free trial if they sign up for a membership.
- Offer casual riders free additional minutes of ride time if they sign up for a membership since their rides tend to be longer.
- Offer casual riders promotions during the summer months which is when their are more casual riders renting bikes.